---
title: "Dubai Houses Price Prediction - ML Pipeline"
author: "Reproducible Research"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```
> [!IMPORTANT]
> **Notice Board: Hardware Constraints**
> Due to the lack of GPU availability in this environment, model training strictly utilizes the CPU. To maintain reasonable execution times, the hyperparameter grids for our models (Random Forest, XGBoost, LightGBM) have been simplified compared to the advanced Python version. While this limits peak model performance slightly, the underlying ML pipeline structure remains fully reproducible and robust.
---

---

# 0. Auto-install & Load Packages

```{r packages}
CRAN <- "https://cloud.r-project.org"

pkgs <- c(
  "arrow", "tidyverse", "skimr", "corrplot", "scales",
  "gridExtra", "caret", "glmnet", "randomForest",
  "xgboost", "lightgbm", "Metrics", "vip",
  "caretEnsemble", "doParallel"
)

to_install <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(to_install)) install.packages(to_install, dependencies = TRUE, repos = CRAN)

suppressPackageStartupMessages({
  library(arrow)
  library(tidyverse)
  library(skimr)
  library(corrplot)
  library(scales)
  library(gridExtra)
  library(caret)
  library(glmnet)
  library(randomForest)
  library(xgboost)
  library(lightgbm)
  library(Metrics)
  library(vip)
  library(caretEnsemble)
  library(doParallel)
})
```

---

# 1. Configuration

```{r config}
# Edit these paths to match your local layout
DATA_DIR   <- "../Data/processed/"   # relative to this .Rmd file
MODELS_DIR <- "models"           # saved model outputs go here

dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

SEED <- 123
set.seed(SEED)
```

---

# 2. Load Data

```{r load-data}
df_train <- read_parquet(file.path(DATA_DIR, "df_train.parquet")) %>% as.data.frame()
df_valid <- read_parquet(file.path(DATA_DIR, "df_valid.parquet")) %>% as.data.frame()
df_test  <- read_parquet(file.path(DATA_DIR, "df_test.parquet")) %>% as.data.frame()

cat("Train rows:", nrow(df_train), "| cols:", ncol(df_train), "\n")
cat("Valid rows:", nrow(df_valid), "| cols:", ncol(df_valid), "\n")
cat("Test  rows:", nrow(df_test),  "| cols:", ncol(df_test),  "\n")
```

---

# 3. Exploratory Data Analysis (EDA)

## 3.1 Quick overview

```{r eda-skim}
skim(df_train)
```

## 3.2 Target distribution

```{r eda-target, fig.height=5}
p1 <- ggplot(df_train, aes(x = price)) +
  geom_histogram(bins = 60, fill = "#3182bd", colour = "white", alpha = .85) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(title = "Price (raw AED)", x = "Price (AED)", y = "Count") +
  theme_minimal()

p2 <- ggplot(df_train, aes(x = price_log1p)) +
  geom_histogram(bins = 60, fill = "#e6550d", colour = "white", alpha = .85) +
  labs(title = "Price (log1p-transformed)", x = "log1p(Price)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## 3.3 Price by property type

```{r eda-type, fig.height=5}
type_cols <- grep("^type_", names(df_train), value = TRUE)

if (length(type_cols) > 0) {
  df_type <- df_train %>%
    select(price_log1p, all_of(type_cols)) %>%
    pivot_longer(all_of(type_cols), names_to = "type", values_to = "flag") %>%
    filter(flag == 1) %>%
    mutate(type = str_remove(type, "^type_"))

  ggplot(df_type, aes(x = reorder(type, price_log1p, median), y = price_log1p)) +
    geom_boxplot(fill = "#74c476", alpha = .7, outlier.size = .5) +
    coord_flip() +
    labs(title = "log1p(Price) by Property Type", x = NULL, y = "log1p(Price)") +
    theme_minimal()
} else {
  message("No 'type_*' columns found - skipping plot.")
}
```

## 3.4 Beds vs Price

```{r eda-beds, fig.height=5}
df_train %>%
  filter(beds <= 10) %>%
  ggplot(aes(x = factor(beds), y = price_log1p)) +
  geom_violin(fill = "#9ecae1", alpha = .7) +
  geom_boxplot(width = .1, outlier.size = .5) +
  labs(title = "log1p(Price) by Number of Beds", x = "Beds", y = "log1p(Price)") +
  theme_minimal()
```

# 3.4b Baths vs Price

```{r eda-baths, fig.height=5}
df_train %>%
  filter(baths <= 10) %>%
  ggplot(aes(x = factor(baths), y = price_log1p)) +
  geom_violin(fill = "#bcbddc", alpha = .7) +
  geom_boxplot(width = .1, outlier.size = .5) +
  labs(title = "log1p(Price) by Number of Baths", x = "Baths", y = "log1p(Price)") +
  theme_minimal()
```

## 3.4c Parking Spaces vs Price

```{r eda-parking, fig.height=5}
df_train %>%
  mutate(parking_capped = ifelse(total_parking_spaces > 5, "6+", as.character(total_parking_spaces))) %>%
  ggplot(aes(x = factor(parking_capped, levels = c(as.character(0:5), "6+")), y = price_log1p)) +
  geom_violin(fill = "#fa9fb5", alpha = .7) +
  geom_boxplot(width = .1, outlier.size = .5) +
  labs(title = "log1p(Price) by Parking Spaces", x = "Parking Spaces", y = "log1p(Price)") +
  theme_minimal()
```

## 3.4d Furnishing Status vs Price

```{r eda-furnish, fig.height=5}
df_train %>%
  mutate(Furnishing = ifelse(furnishing_Unfurnished == TRUE, "Unfurnished", "Furnished")) %>%
  ggplot(aes(x = Furnishing, y = price_log1p, fill = Furnishing)) +
  geom_violin(alpha = .7, show.legend = FALSE) +
  geom_boxplot(width = .1, outlier.size = .5, show.legend = FALSE) +
  scale_fill_manual(values = c("Furnished" = "#a1d99b", "Unfurnished" = "#bdbdbd")) +
  labs(title = "log1p(Price) by Furnishing Status", x = "Status", y = "log1p(Price)") +
  theme_minimal()
```

## 3.4e Completion Status vs Price

```{r eda-completion, fig.height=5}
df_train %>%
  mutate(Status = ifelse(completion_status_Ready == TRUE, "Ready", "Off-Plan")) %>%
  ggplot(aes(x = Status, y = price_log1p, fill = Status)) +
  geom_violin(alpha = .7, show.legend = FALSE) +
  geom_boxplot(width = .1, outlier.size = .5, show.legend = FALSE) +
  scale_fill_manual(values = c("Ready" = "#9ecae1", "Off-Plan" = "#fdae6b")) +
  labs(title = "log1p(Price) by Completion Status", x = "Status", y = "log1p(Price)") +
  theme_minimal()
```

## 3.4f Quarter vs Price

```{r eda-quarter, fig.height=5}
df_train %>%
  ggplot(aes(x = factor(quarter), y = price_log1p)) +
  geom_violin(fill = "#dadaeb", alpha = .7) +
  geom_boxplot(width = .1, outlier.size = .5) +
  labs(title = "log1p(Price) by Transaction Quarter", x = "Quarter", y = "log1p(Price)") +
  theme_minimal()
```

## 3.5 Geographic spread

```{r eda-geo, fig.height=5}
if (all(c("Longitude", "Latitude") %in% names(df_train))) {
  ggplot(df_train, aes(x = Longitude, y = Latitude, colour = price_log1p)) +
    geom_point(size = .6, alpha = .5) +
    scale_colour_viridis_c(option = "plasma", name = "log1p(price)") +
    labs(title = "Property Locations coloured by Price") +
    theme_minimal()
}
```

## 3.6 Correlation heatmap

```{r eda-corr, fig.height=7, fig.width=9}
num_df <- df_train %>% select(where(is.numeric))
num_df <- num_df[, apply(num_df, 2, var, na.rm = TRUE) > 0]
cor_mat <- cor(num_df, use = "pairwise.complete.obs")

corrplot(cor_mat,
         method = "color", type = "upper",
         tl.cex = 0.55, tl.col = "black",
         col    = colorRampPalette(c("#2166ac", "white", "#d6604d"))(200),
         title  = "Pearson Correlation Matrix",
         mar    = c(0, 0, 1.5, 0))
```

## 3.7 Top features correlated with target

```{r eda-top-corr}
cor_mat["price_log1p", ] %>%
  sort(decreasing = TRUE) %>%
  enframe(name = "feature", value = "correlation") %>%
  filter(!feature %in% c("price_log1p", "price")) %>%
  slice_head(n = 20) %>%
  ggplot(aes(x = reorder(feature, correlation), y = correlation)) +
  geom_col(fill = "#6baed6") +
  coord_flip() +
  labs(title = "Top 20 Features Correlated with log1p(Price)",
       x = NULL, y = "Pearson r") +
  theme_minimal()
```

---

# 4. Feature Matrix Setup

```{r features}
drop_cols <- c("price", "price_log1p", "__index_level_0__")

X_train <- df_train %>% select(-any_of(drop_cols))
names(X_train) <- make.names(names(X_train))
y_train <- df_train$price_log1p

X_valid <- df_valid %>% select(-any_of(drop_cols))
names(X_valid) <- make.names(names(X_valid))
y_valid <- df_valid$price_log1p

X_test  <- df_test %>% select(-any_of(drop_cols))
names(X_test) <- make.names(names(X_test))
y_test  <- df_test$price_log1p

cat("Feature count:", ncol(X_train), "\n")
```

---

# 5. Cross-Validation & Helpers

```{r cv-setup}
cl <- makePSOCKcluster(max(1, parallel::detectCores() - 1))
registerDoParallel(cl)

cv_ctrl <- trainControl(
  method          = "none",
  allowParallel   = TRUE,
  verboseIter     = FALSE
)

eval_model <- function(model, X, y_true, model_name = "") {
  pred_log <- predict(model, newdata = X)
  pred_raw <- expm1(pred_log)
  y_raw    <- expm1(y_true)
  tibble(
    Model    = model_name,
    MAE_log  = round(mae(y_true,  pred_log), 5),
    RMSE_log = round(rmse(y_true, pred_log), 5),
    R2_log   = round(cor(y_true,  pred_log)^2, 4),
    MAE_AED  = round(mae(y_raw,   pred_raw), 0),
    RMSE_AED = round(rmse(y_raw,  pred_raw), 0)
  )
}

results <- list()
```

---

# 6. Simple Baseline Models

## 6.1 OLS (Linear Regression)

```{r model-ols}
set.seed(SEED)
model_ols <- train(
  x          = X_train,
  y          = y_train,
  method     = "lm",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale")
)
print(model_ols)
results[["OLS"]] <- eval_model(model_ols, X_valid, y_valid, "OLS")
saveRDS(model_ols, file.path(MODELS_DIR, "model_ols.rds"))
cat("OLS saved.\n")
```

## 6.2 Ridge Regression

```{r model-ridge}
set.seed(SEED)
model_ridge <- train(
  x          = X_train,
  y          = y_train,
  method     = "glmnet",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  tuneGrid   = data.frame(alpha = 0, lambda = 0.1)
)
results[["Ridge"]] <- eval_model(model_ridge, X_valid, y_valid, "Ridge")
saveRDS(model_ridge, file.path(MODELS_DIR, "model_ridge.rds"))
cat("Ridge saved.\n")
```

## 6.3 ElasticNet

```{r model-enet}
set.seed(SEED)
model_enet <- train(
  x          = X_train,
  y          = y_train,
  method     = "glmnet",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  tuneGrid   = data.frame(alpha = 0.5, lambda = 0.1)
)
results[["ElasticNet"]] <- eval_model(model_enet, X_valid, y_valid, "ElasticNet")
saveRDS(model_enet, file.path(MODELS_DIR, "model_enet.rds"))
cat("ElasticNet saved.\n")
```

---

# 7. Intermediate Model

## 7.1 Random Forest

```{r model-rf}
set.seed(SEED)
model_rf <- train(
  x          = X_train,
  y          = y_train,
  method     = "rf",
  trControl  = cv_ctrl,
  tuneGrid   = data.frame(mtry = 10),
  ntree      = 50,
  importance = TRUE
)
vip(model_rf, num_features = 15, aesthetics = list(fill = "#41ab5d")) +
  labs(title = "Random Forest - Variable Importance") + theme_minimal()
results[["RandomForest"]] <- eval_model(model_rf, X_valid, y_valid, "RandomForest")
saveRDS(model_rf, file.path(MODELS_DIR, "model_rf.rds"))
cat("Random Forest saved.\n")
```

---

# 8. Gradient Boosting Models

## 8.1 XGBoost

```{r model-xgb}
set.seed(SEED)
dtrain_xgb <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)

xgb_params <- list(
  objective          = "reg:squarederror",
  max_depth          = 6,
  eta                = 0.10,
  gamma              = 0,
  colsample_bytree   = 0.7,
  min_child_weight   = 5,
  subsample          = 0.8
)

model_xgb <- xgb.train(
  params  = xgb_params,
  data    = dtrain_xgb,
  nrounds = 100,
  verbose = 0
)

pred_xgb_valid <- predict(model_xgb, as.matrix(X_valid))
results[["XGBoost"]] <- tibble(
  Model    = "XGBoost",
  MAE_log  = round(mae(y_valid,         pred_xgb_valid), 5),
  RMSE_log = round(rmse(y_valid,        pred_xgb_valid), 5),
  R2_log   = round(cor(y_valid,         pred_xgb_valid)^2, 4),
  MAE_AED  = round(mae(expm1(y_valid),  expm1(pred_xgb_valid)), 0),
  RMSE_AED = round(rmse(expm1(y_valid), expm1(pred_xgb_valid)), 0)
)

xgb.importance(model = model_xgb) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#fd8d3c") + coord_flip() +
  labs(title = "XGBoost - Variable Importance", x = NULL, y = "Gain") +
  theme_minimal()

xgb.save(model_xgb, file.path(MODELS_DIR, "model_xgb.model"))
cat("XGBoost saved.\n")
```

## 8.2 LightGBM

```{r model-lgbm}
set.seed(SEED)
dtrain_lgbm <- lgb.Dataset(data = as.matrix(X_train), label = y_train)

lgbm_params <- list(
  objective         = "regression",
  metric            = "mae",
  learning_rate     = 0.05,
  num_leaves        = 127,
  min_child_samples = 20,
  feature_fraction  = 0.7,
  bagging_fraction  = 0.8,
  bagging_freq      = 5,
  reg_alpha         = 0.1,
  reg_lambda        = 1.0,
  verbose           = -1
)

model_lgbm <- lgb.train(
  params  = lgbm_params,
  data    = dtrain_lgbm,
  nrounds = 100,
  verbose = -1
)

pred_lgbm_valid <- predict(model_lgbm, as.matrix(X_valid))
results[["LightGBM"]] <- tibble(
  Model    = "LightGBM",
  MAE_log  = round(mae(y_valid,         pred_lgbm_valid), 5),
  RMSE_log = round(rmse(y_valid,        pred_lgbm_valid), 5),
  R2_log   = round(cor(y_valid,         pred_lgbm_valid)^2, 4),
  MAE_AED  = round(mae(expm1(y_valid),  expm1(pred_lgbm_valid)), 0),
  RMSE_AED = round(rmse(expm1(y_valid), expm1(pred_lgbm_valid)), 0)
)

lgb.importance(model_lgbm) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#9e9ac8") + coord_flip() +
  labs(title = "LightGBM - Variable Importance (Gain)", x = NULL, y = "Gain") +
  theme_minimal()

lgb.save(model_lgbm, file.path(MODELS_DIR, "model_lgbm.txt"))
cat("LightGBM saved.\n")
```

---

# 9. Stacking Ensemble

```{r model-stacking}
set.seed(SEED)

cv_ctrl_idx <- trainControl(
  method          = "cv",
  number          = 2,
  savePredictions = "final",
  allowParallel   = TRUE,
  index           = createFolds(y_train, k = 2, returnTrain = TRUE)
)

model_list <- caretList(
  x         = X_train,
  y         = y_train,
  trControl = cv_ctrl_idx,
  tuneList  = list(
    lm    = caretModelSpec(method = "lm"),
    ridge = caretModelSpec(method = "glmnet",
                           tuneGrid = data.frame(alpha = 0, lambda = 0.1)),
    rf    = caretModelSpec(method = "rf",
                           tuneGrid = data.frame(mtry = 10), ntree = 50)
  )
)

model_stack <- caretStack(
  model_list,
  method    = "glmnet",
  trControl = trainControl(method = "none"),
  tuneGrid  = data.frame(alpha = 0, lambda = 0.1)
)

print(model_stack)
results[["Stacking"]] <- eval_model(model_stack, X_valid, y_valid, "Stacking")
saveRDS(model_stack, file.path(MODELS_DIR, "model_stacking.rds"))
cat("Stacking ensemble saved.\n")
```

---

# 10. Model Comparison

```{r comparison}
stopCluster(cl)
registerDoSEQ()

results_df <- bind_rows(results) %>% 
  mutate(
    MAE_log  = as.numeric(MAE_log),
    RMSE_log = as.numeric(RMSE_log),
    R2_log   = as.numeric(R2_log),
    MAE_AED  = as.numeric(MAE_AED),
    RMSE_AED = as.numeric(RMSE_AED)
  ) %>%
  arrange(MAE_log)
knitr::kable(results_df, caption = "Model Comparison on Validation Set")
```

```{r comparison-plot}
results_df %>%
  arrange(MAE_log) %>%
  mutate(Model = fct_inorder(Model)) %>%
  ggplot(aes(x = Model, y = MAE_log, fill = Model)) +
  geom_col(show.legend = FALSE, alpha = .85) +
  geom_text(aes(label = round(MAE_log, 4)), hjust = -.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Validation MAE (log1p scale) - Lower is Better",
       x = NULL, y = "MAE (log1p)") +
  theme_minimal()
```

```{r comparison-aed}
results_df %>%
  arrange(MAE_AED) %>%
  mutate(Model = fct_inorder(Model)) %>%
  ggplot(aes(x = Model, y = MAE_AED / 1e6, fill = Model)) +
  geom_col(show.legend = FALSE, alpha = .85) +
  geom_text(aes(label = round(MAE_AED / 1e6, 2)), hjust = -.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Validation MAE (AED millions) - Lower is Better",
       x = NULL, y = "MAE (millions AED)") +
  theme_minimal()
```

---

# 11. Best Model - Test Set Evaluation

```{r test-eval}
best_model_name <- results_df$Model[1]
cat("Best model on validation:", best_model_name, "\n")

if (best_model_name == "LightGBM") {
  pred_test_log <- predict(model_lgbm, as.matrix(X_test))
} else if (best_model_name == "XGBoost") {
  pred_test_log <- predict(model_xgb, as.matrix(X_test))
} else {
  best_model <- switch(best_model_name,
    "OLS"          = model_ols,
    "Ridge"        = model_ridge,
    "ElasticNet"   = model_enet,
    "RandomForest" = model_rf,
    "Stacking"     = model_stack
  )
  pred_test_log <- predict(best_model, newdata = X_test)
}

pred_test_raw <- expm1(pred_test_log)
y_test_raw    <- expm1(y_test)

cat(sprintf("Test MAE  (log):  %.5f\n",  mae(y_test,     pred_test_log)))
cat(sprintf("Test RMSE (log):  %.5f\n", rmse(y_test,     pred_test_log)))
cat(sprintf("Test R2   (log):  %.4f\n",  cor(y_test,     pred_test_log)^2))
cat(sprintf("Test MAE  (AED):  %.0f\n",  mae(y_test_raw, pred_test_raw)))
```

```{r pred-vs-actual, fig.height=5}
tibble(actual = y_test, predicted = pred_test_log) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = .25, size = .8, colour = "#2171b5") +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 1) +
  labs(title    = paste("Predicted vs Actual -", best_model_name, "(Test Set)"),
       subtitle = "Red line = perfect prediction",
       x = "Actual log1p(Price)", y = "Predicted log1p(Price)") +
  theme_minimal()
```

```{r residuals, fig.height=5}
tibble(fitted = pred_test_log, residuals = y_test - pred_test_log) %>%
  ggplot(aes(x = fitted, y = residuals)) +
  geom_point(alpha = .25, size = .8, colour = "#e6550d") +
  geom_hline(yintercept = 0, colour = "black", linewidth = .8) +
  labs(title = "Residuals vs Fitted (Test Set)",
       x = "Fitted values", y = "Residuals") +
  theme_minimal()
```

---

# 12. How to Reload Saved Models

```{r reload-demo, eval=FALSE}
# caret / caretEnsemble models (.rds)
model_ols_loaded   <- readRDS(file.path(MODELS_DIR, "model_ols.rds"))
model_ridge_loaded <- readRDS(file.path(MODELS_DIR, "model_ridge.rds"))
model_enet_loaded  <- readRDS(file.path(MODELS_DIR, "model_enet.rds"))
model_rf_loaded    <- readRDS(file.path(MODELS_DIR, "model_rf.rds"))
model_stack_loaded <- readRDS(file.path(MODELS_DIR, "model_stacking.rds"))

# XGBoost (.model)
model_xgb_loaded <- xgb.load(file.path(MODELS_DIR, "model_xgb.model"))
preds_xgb <- predict(model_xgb_loaded, as.matrix(X_test))

# LightGBM (.txt)
model_lgbm_loaded <- lgb.load(file.path(MODELS_DIR, "model_lgbm.txt"))
preds_lgbm <- predict(model_lgbm_loaded, as.matrix(X_test))
```

---

# 13. Session Info

```{r session}
sessionInfo()
```
